Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies

### 2024-11-09 manuscript/manuscript.qmd
Mega additions and edits and reorganization

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: 2024-11-08
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
epigraph:
  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
keywords:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true
---

@@

::: {.callout-note}
# ARNOVA outline

Cleaned up, simplified outline that I'll expand and integrate with the rest of the outline/draft below
:::

@@

# The fundamental problem of casual inference

Explain that here…

# Visualizing, identifying, and isolating causal relationships with DAGs

# Isolating causal relationships

# Situation-based approaches

## RCTs

## Diff-in-diff + TWFE

## RDD

## IV

# Adjustment-based approaches

## Regression adjustment

## IPW and g-computation

# Hybrid approaches

@@

@@

# Why is causal inference so difficult?

@@

# A vocabulary for observational causal inference (ANDREW)

In the absence of random assignment, finding causal effects in observational data where units of analysis can self select their treatments and outcomes thus appears impossible.

Goal of much research in recent years has been to create methods that allow researchers to make plausible causal arguments using observed, non-experimental data. Judea Pearl, computer science, and epidemiology provide us with a special framework for observational causal inference. We introduce the vocabulary and grammar of this framework below.

## Causal diagrams

[@PearlMackenzie:2018; @MorganWinship:2014; @PearlGlymourJewell:2016]

DAGs encode our understanding of the data generating process - what in the natural and social world leads to the treatment/exposure, the outcome, and both simultaneously. It represents a type of philosophical model of how the observed world works. Importantly, causal graphs also encode our assumptions of how the world works.

- Nodes - nodes can be unmeasurable, or even unobserved
- Edges - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two.
- Acyclicity - can't get back to a node - represents flow of time - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1

confounder <- ggplot(data_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_confounder, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Confounder", subtitle = "(Fork)") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange),
        plot.subtitle = element_text(hjust = 0.5, color = clrs$orange))

mediator <- ggplot(data_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_mediator, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Mediator", subtitle = "(Chain)") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple),
        plot.subtitle = element_text(hjust = 0.5, color = clrs$purple))

collider <- ggplot(data_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_collider, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Collider", subtitle = "(Inverted fork)") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$red),
        plot.subtitle = element_text(hjust = 0.5, color = clrs$red))

## Causal identification

@@

# Adjustment-based identification (MENG AND ANDREW)

TODO: Use simulated data based on a DAG that has a mediator, collider, and a bunch of confounders

TODO: Replicate a study + draw a DAG for it + use IPW to close back doors

## Backdoor adjustment

@@

There are many different ways—regression, matching, IPW—each with different nuances.

### Regression

### Matching

### IPW

## Illustration

### Simulated example

### Real example

## Frontdoor adjustment

A less common approach to removing confounding is frontdoor adjustment

## Other methods?

# Design-based identification (ANDREW AND MENG)

@@

The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects. Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs (see @fig-design-based-dags))

@@

rct <- ggplot(data_rct, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = status), size = 12) +
  geom_dag_text(size = 4) +
  annotate(geom = "text", x = 1, y = 1.2, label = "X = x") +
  scale_color_manual(values = c(clrs$green, clrs$blue), 
                     na.value = clrs$yellow, guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Randomized trial", 
       subtitle = "Randomization deletes all arrows into X") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

@@

For each of these:

1. Sort explanation of what they require, how they work
2. Explanation of how that connects to the DAG and how the graph reveals the identification strategy
3. Point to some nonprofit-related papers that use these

Quasi-experiments (design-based—special situations)

- Diff-in-diff
- RDD
- IV

## Experiments

@@

## Diff-in-diff

@@

## Regression discontinuity

@@

## Instrumental variables

@@

# Tools for researchers and practitioners (BOTH)

@@

## Dealing with time

- Epidemiology and polsci and marginal structural models
- Economics and TWFE and synthetic controls and event studies - fancier diff-in-diff

# Conclusion (BOTH)

@@

>>>>

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: 2024-11-08
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
epigraph:
  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
keywords:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true

params:
  wordcount: |
    <strong>{{< words-sum body-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references
---

::: {.content-visible when-meta="wordcount-banner"}
{{< include _extensions/andrewheiss/wordcount/banner.html >}}
:::

@@

@@

@@

Stuff from first version of introduction:

@@

Textbook treatments of causal inference approahces are typically abstract and generic, explaining how treatment $X$ causes outcome $Y$ after adjusting for confounders $Z$. To make these principles as concrete as possible, we use a practical running example relevant to nonprofit education and practice. Experiential philanthropy - Newark, BYU Grantwell and Be a Philanthropist. These programs provide students interested in the nonprofit sector with hands on, practical experience with philanthropy. Beyond their pedagogical aims, these programs are designed to instill a long-term interest and participation in philanthropy.

> Do experiential philanthropy programs cause long-term charitable or prosocial behavior?

We do not provide any evidence to answer this question directly. Rather, we use simulated data to explore multiple hypothetical research designs that could be used to quantititavely measure the causal impact of these programs. These causal inference approaches can be applied to a wide range of nonprofit-focued questions, though some methods are better suited than others, depending on the context of the research question. TODO We provide examples of possible nonprofit-related causal inquiries that are amenable to each method

# A visual vocabulary for causal inference

## Causation through listening

Judea Pearl, computer science, and epidemiology blah blah blah provide us with a specialized framework for describing causal theories and making causal inferences.

At the core of causal inquiry is the notion of "causation," or what it means when we say that an intervention causes an outcome. In Pearl's framework, causation can be defined using a metaphor of listening and responding:

> A variable $X$ is a cause of a variable $Y$ if $Y$ in any way relies on $X$ for its value.… $X$ is a cause of $Y$ if $Y$ listens to $X$ and decides its value in response to what it hears. [@PearlGlymourJewell:2016, 5--6].

In the context of our example causal question, we posit that prosocial behavior listens to, responds to, or is caused by experiential philanthropy. We have a theoretical reason to believe that the intervention and the outcome are associated with each other in a sequential, nonspurious way—it is reasonable to assume that having hands-on experience with philanthropy today would influence future attitudes toward charitable giving. Specific educational interventions are not the only cause of prosocial behavior. This behavior listens and responds to other factors, including family background, income, education, and personal attitudes and opinions towards volunteerism and charity—experiential philanthropy is just one of many causes.

## Directed acyclic graphs (DAGs)

We can formally represent our theory of how an intervention ($X$) causes an outcome ($Y$) using a directed acyclic graph, or DAG [@PearlMackenzie:2018; @PearlGlymourJewell:2016; @MorganWinship:2014]. DAGs encode our understanding of the data generating process, or which phenomena cause the treatment, the outcome, and both simultaneously. These causal graphs are a philosophical model of the associations between different phenomena and our assumptions behind those relationships.

- **Nodes** - nodes can be unmeasurable, or even unobserved
- **Edges** - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two. Unlike other graphical approaches like structural equation models (SEMs) that assume all relationships are additive and linear, DAG arrows are nonparametric and can represent any kind of functional form (linear, multiplicative, polynomial, exponential, etc.). Accordingly, when we say that $X \rightarrow Y$, we mean that $X$ causally affects $Y$ in some general way, not necessarily linearly [@Rohrer:2018].
- **Acyclicity** - can't get back to a node - represents flow of time, or temporal ordering - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1

```{r}
#| label: fig-dag-basic-real
#| fig-width: 4.5
#| fig-height: 2.75
#| out-width: 60%
#| fig-cap: "An example directed acyclic graph that shows how experiential philanthropy causes charitable behavior, while incomes causes both"
#| include: true

data_confounder_example <- dagify(
  Y ~ Z + X,
  X ~ Z,
  coords = list(x = c(X = 1, Y = 3, Z = 2),
                y = c(X = 1, Y = 1, Z = 2)),
  labels = c(X = "Experiential\nphilanthropy", Y = "Charitable\nbehavior", Z = "Income")
) |> 
  tidy_dagitty()

ggplot(data_confounder_example, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_width = 1) +
  geom_dag_point(aes(color = name), size = 12) +
  geom_dag_text(aes(label = name)) +
  geom_dag_text(
    data = filter(data_confounder_example, name != "Z"), aes(label = label), 
    size = 4, nudge_y = -0.32, color = "black", lineheight = 1
  ) +
  geom_dag_text(
    data = filter(data_confounder_example, name == "Z"), aes(label = label), 
    size = 4, nudge_y = 0.25, color = "black", lineheight = 1
  ) +
  scale_color_manual(values = c("X" = clrs$green, "Y" = clrs$blue, "Z" = clrs$yellow), guide = "none") +
  coord_cartesian(xlim = c(0.75, 3.25), ylim = c(0.6, 2.2)) +
  theme_np_dag()
```

Importantly, DAGs do more than simply represent the relationships and associations between variables. They provide us with ways to isolate specific relationships of interest. TODO: Rohrer, 28: Even if a researcher does not rely fully on a DAG, mapping out the underlying data-generating process

## Statistical associations and causal structures

confounder <- ggplot(data_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_confounder, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Confounder") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange))

mediator <- ggplot(data_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_mediator, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Mediator") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple))

collider <- ggplot(data_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = node_color), size = 12) +
  geom_dag_text(data = filter(data_collider, name != "Z"), size = 4) +
  geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Collider") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5, color = clrs$red))

Even if we do not rely on a DAG for complete identification of causal effects, mapping out the relationships between nodes in the data-generating process is a useful exercise in any kind of quantitative, regression-based analysis, including non-causal descriptive work. Researchers often take a "kitchen sink" approach to statistical control, including all potential covariates in a model. Understanding the underlying DAG provides clearer guidance about what to control for, helping researchers avoid "bad controls" [@cite]. For instance, if a potential covariate is only associated with the outcome and has no relationship to the treatment, it is not a confounder and need not be included as a control variable in a model, since it can reduce the precision of the model estimates. If a potential covariate is a collider, caused by both the treatment and the outcome, the model results will be biased or distorted.

# Identifying and isolating causal relationships

@@

## Causal identification and *d*-separation

@@

@@

## Circumstantial identification

@@

The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects. Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs [see @fig-design-based-dags]

@@

rct <- ggplot(data_rct, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
  geom_dag_point(aes(color = status), size = 12) +
  geom_dag_text(size = 4) +
  annotate(geom = "text", x = 1, y = 1.2, label = "X = x") +
  scale_color_manual(values = c(clrs$green, clrs$blue), 
                     na.value = clrs$yellow, guide = "none") +
  coord_cartesian(ylim = c(0.95, 2.05)) +
  labs(title = "Randomization", 
       subtitle = "Randomization deletes all arrows into X") +
  theme_np_dag() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

@@

### RCTs

@@

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it - survey vignette experiments, conjoint experiments, field experiments, other kinds of RCTs
- Possible nonprofit research questions that could use it

### Diff-in-diff + TWFE

@@

### RDD

@@

### IV

@@

These used to be common, but have become less popular because of the difficulty in finding a valid instrument that meets all the criteria for clean identification. @rainfall-instrument-paper. Randomized promotion works well, though - @worldbank-book

## Adjustment-based identification

### Regression adjustment

### IPW and g-computation

Despite its popularity in epidemiology and public health, to our knowledge, there are no studies in NVSQ, Voluntas, or NML that employ inverse probability weighting for covariate adjustment for causal inference. There are X that use matching, and there are many that use regression-based adjustment in explicitly non-causal ways (but Haber et al paper finds that these things imply causation anyway). For instance, Huafang et al. paper explicity talks about associations after controlling for ostensible confounders, but also uses causal language like "program impacts" - C-word article and not shying away from causation

## Hybrid approaches

@@

@@

We believe that adjustment-based approaches like regression adjustment, matching, and IPW are highly amenable to nonprofit administrative data. Finding specific situations like diff-in-diff is popular in public policy analysis and econometrics because the scale of the data is much larger—researchers can analyze the effect of state- or national-level policies in difference-in-difference designs, or observe the behavior of millions of participants in programs like Medicaid or the Affordable Care Act for regression discontinuity designs. A smaller nonprofit interested in evaluating the impact of a new community program is most likely unable to consider these types of situation-specific quasi-experimental designs to isolate the causal effect. However, with careful theoretical thinking; the development of a robust DAG that describes potential confounders, mediators, and colliders; the use of high quality administrative data; and intentional sensitivity analysis that probes the strength of possible causal relationships when faced with unmeasured confounding [@that-sensmakr-paper; @lucys-tipr-paper]; nonprofit researchers can tell plausible causal stories.

# References
